Hydrous melting and partitioning in and above the mantle transition zone: insights from water-rich MgO-SiO$_2$-H$_2$O experiments

### 2016-05-12 revised_submission/manuscript/MSH.tex
Submitted

%% Group authors per affiliation:
\author{R. Myhill}
\address{Bayerisches Geoinstitut, Universit\"{a}t Bayreuth, Universit\"{a}tsstrasse 30, 95447 Bayreuth, Germany \\ Now at School of Earth Sciences, University of Bristol, Wills Memorial Building, Queens Road, Bristol BS8 1RJ, United Kingdom}
\cortext[mycorrespondingauthor]{Corresponding author: R. Myhill}
\ead{myhill.bob@gmail.com}

@@

Constraining melt compositions remains a major challenge for the field of high-pressure experimental petrology. The primary problem is that water-rich melts are unquenchable. During quenching, some of the water in the melt crystallises as hydrous minerals such as brucite, but a significant proportion crystallises as water ice at high pressure, which is then lost as water vapour during decompression and subsequent sample preparation. The original H$_2$O contents of the partial hydrous melts are normally estimated by mass balance and/or by deficits in microprobe totals \citep[e.g.][]{OMY2000, DDFK2005, LSOK2011, MSUP2007}. Using mass balance to calculate H$_2$O contents depends on being able to estimate the volume proportions of solid and quench phases, recognise and subtract quench overgrowths where they exist, and calculate the effective ambient-pressure density of the melt. Such estimates are typically associated with significant uncertainties. Deficits in microprobe totals are sometimes reasonable for estimating water contents in solid phases, provided the microprobe standards are suitable, and the microprobe is exceptionally well-calibrated (even a small deficit implies a large water content). However, for quench melts the deficits are attributable to water held in quench phases \emph{and} porosity within the volume sampled by the defocused beam. For this reason, defocused beam techniques should probably be restricted to estimating non-volatile element ratios (such as Mg:Si).

Here we use the approach of bracketing the liquidus surface to determine melt composition as a function of temperature in the system MgO-SiO$_2$-H$_2$O (MSH). This technique requires knowledge of the starting composition, techniques to effectively minimise water loss from the capsule and an ability to recognise the difference between liquidus and quench phases. We use this technique to develop thermodynamic models for water solubility in melts, and apply these to models of water partitioning between mantle phases and melts. We then apply these models to better understand hydrous melting under transition zone conditions.
%XXXX what, in general, have you done in this paper?

\section{Experimental and analytical methods}
Starting compositions were created from a mixture of high-purity brucite, quartz and platinic acid (hexahydroxyplatinate(IV); H$_2$Pt(OH)$_6$). We use platinic acid to increase water contents to those expected of hydrous melts in equilibrium with mantle phases under subduction/convecting mantle conditions. Platinic acid is a pale yellow compound which is stable to about 130 $^{\circ}$C \citep{Nagano2002}. Its relatively high stability, high water content and the inert nature of the breakdown product (in a system where redox reactions are absent) make it an excellent water source for high pressure hydrous melting experiments. Before weighing, quartz was dried at 1000 $^{\circ}$C for 12 hours, while brucite was heated to 250 $^{\circ}$C, also over 12 hours. Both powders were stored in a desiccating oven at 130 $^{\circ}$C. Platinic acid is hygroscopic, so was stored in a vacuum desiccator. Powders were weighed and ground dry in an agate mortar for 30 minutes, using a mask, goggles and gloves to avoid physical contact with the platinic acid. Prepared powders were stored in glass vials in a vacuum desiccator until use.

@@

\begin{figure}[ht!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/periclase}
  \caption{The periclase-water phase diagram at 13 GPa, based on current experimental results. Model fits to the data use the melt formulation of \cite{SS1985}, with r=1 and K=K(T) (see discussion in main text). Properties of the anhydrous solid and melt are taken from published literature \citep{SLB2011, DKS2013}. An additional subregular interaction term is required to match the composition and temperature of brucite dehydration.}
  \label{fig:MH}
\end{figure}

@@

\begin{figure}[ht!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/forsterite}
      \caption{The forsterite-water phase diagram at 13 GPa, based on current experimental results. The thermodynamics of pure forsterite are taken from the literature \citep{SLB2011}, and the properties of the anhydrous melt are from the mixing model described in the text. At high temperatures, the forsterite liquidus is metastable with respect to periclase + melt. High temperature data come from the study of \cite{NDMPMF2016}.}
  \label{fig:foH}
\end{figure}
\clearpage

@@

\begin{figure}[ht!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/enstatite}
  \caption{The clinoenstatite-water phase diagram at 13 GPa, based on current experimental results. The thermodynamics of pure high pressure clinoenstatite are taken from the literature \citep{SLB2011}, and the properties of the anhydrous melt are from the mixing model described in the text. High temperature data come from the study of \cite{NDMPMF2016}.}
  \label{fig:eoH}
\end{figure}

@@

\begin{figure}[ht!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/stishovite}
      \caption{The silica-water phase diagram at 13 GPa, based on current experimental results. The thermodynamics of stishovite and the anhydrous melt are described in the text. At high temperatures, the stishovite liquidus is metastable with respect to coesite and melt.}
  \label{fig:SH}
\end{figure}

@@

%\citep{SU2001, KOM2005, AA1980, Wunder1998, AFRH2001, Nagano2002, MFY2002, Sta2004, MFY2004, SUTG2001, NM2008, DM2010, ZK1998}

@@

\begin{figure}[ht!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/experimental-ternary}
  \caption{A preliminary ternary liquidus diagram at 13 GPa, based on current experimental results. Contours are in $^{\circ}$C, and mineral abbreviations are liquidus fields as follows: per - periclase, fo - forsterite, en - clinoenstatite, stv - stishovite, coe - coesite, phB - phase B, br - brucite, chond - hydroxychondrodite, ice - ice VII. The locations of the contours are constrained by observed liquidus phases and the liquidus-bracketing results presented here. Previous work on hydrous phase stability and Mg:Si ratios for cotectic melts are also included \citep{MSUP2007}.}
  \label{fig:ternary}
\end{figure}

@@

% Densities
%\cite{MF2010}
%\cite{BK2003}
%\cite{MSK2008}

%So at least in my models there are three quantities that are important for the pattern of melt migration, the permeability, the compaction viscosity and the shear viscosity (if you have elasticity/plasticity, other things come in, too). Everywhere in the equations where you have the permeability, it's divided by the melt viscosity, and the permeability itself is normally assumed to scale with phi^3 (or less common phi^2), and at least I didn't see any parametrisations describing the effect of temperature on the permeability, so I would assume it's small. So for the permeability, you can just use the values you get for your different P-T-X conditions, and see if the overall term permeability/melt_viscosity increases or decreases.

%For the viscosities of the solid you have the influence of temperature on the viscosity (which I think is similar for shear and compaction viscosity) and you have the effect of porosity, eta ~ exp(alpa * phi) with phi~-27 for the shear viscosity (I think this is what Rich Katz uses in his models, but I would have to look it up to be sure), and xi ~ phi0/phi for the compaction viscosity, and you can do the same here, compare how they change for your different P-T-X conditions.

@@

Partitioning of hydrogen between mineral and melt phases can be described using a partition coefficient D:
\begin{equation}
  D^{a/b}_{H_2O} = c^a_{H_2O} / c^b_{H_2O}
\end{equation}
where $c^i_{H_2O}$ is the concentration of H$_2$O (in wt\%) in phase $i$ \citep{KB2006}. In quenchable phases, there are a number of ways that concentration can be measured, including FTIR, SIMS and ERDA. In the case of melts, which are unquenchable, $c^{melt}_{H_2O}$ in partitioning studies has most commonly been estimated by mass balance or by assuming that the deficit in microprobe totals is due to the fraction of oxygen bonded to hydrogen. In two studies on Fe-free wadsleyite, \cite{DDFK2005} and \cite{LSOK2011} present estimates of water content in melts calculated from microprobe deficits. A similar study on ringwoodite was undertaken by \cite{OMY2000}. The T-X$_{H_2O}$ dependencies of the melts in equilibrium with wadsleyite or ringwoodite are strikingly different (Figure \ref{fig:fo_wad_rw_melt}). For example, melt in equilibrium with wadsleyite at 15 GPa and 1400 $^{\circ}$C was estimated to contain 10.6-13.3 wt\% H$_2$O, while melt compositions in equilibrium with ringwoodite at 20 GPa and 1400 $^{\circ}$C suggest water contents of 57-66 wt\% H$_2$O. Modelled Mg$_2$SiO$_4$ melting temperatures increase by only 40 K between 15 and 20 GPa, and entropies of melting at these two pressures are not very different (92.4 vs. 97 J/K/mol), so unless the thermodynamics of mixing in the liquid change markedly between these pressures, estimates of $c^{melt}_{H_2O}$ for the ringwoodite or wadsleyite studies (or both) must be somewhat inaccurate. We note that two of the three melt compositions in equilibrium with ringwoodite fall above the temperatures indicating that hydrogen exists only as molecular H$_2$O \citep{SS1985}. In addition, the calculated liquidus temperatures in the wadsleyite studies at 15 GPa are lower than those reported for olivine at 5.5 GPa \citep{Inoue1994}. Both of these observations, therefore, raise inconsistencies which are difficult to address within a self-consistent thermodynamic model.

@@

\begin{figure}[ht!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/fo_wad_rw}
  \caption{Water concentrations in wadsleyite, ringwoodite and melt, modelled at 15 and 20 GPa. Experimental data is taken from three published studies \citep{OMY2000,DDFK2005,LSOK2011}.}
  \label{fig:fo_wad_rw_melt}
\end{figure}

@@

\begin{table}[]
\centering
\caption{Thermodynamic models for water solubility in forsterite, wadsleyite and ringwoodite. The forsterite model is taken from the literature \cite{ZGK2004}.}
\label{table:partitioning}
\begin{tabular}{lllll}
Phase & $P_{\textit{\tiny{ref}}}$ (GPa) & A [$10^{-12}$ kg/(kgPa)] & $\Delta H_{P_{\textit{\tiny{ref}}}}$ [kJ/mol] & $\Delta V$ [cm$^3$/mol] \\
\hline
% Olivine & 1.2e10 & 1.74e-11 & 1.70e5 & 1e-5 \\ Not a bad fit for the Zhao enthalpy
% Olivine & 1.2e10 & 5e-12 & 1.70e5 & 1e-5 \\ Zhao et al., 2004, water contents much too high
Olivine & 12 & 13.5 & 170 & 10 \\
Wadsleyite & 15 & 1.8 & 160 & 10 \\
Ringwoodite & 20 & 1.2 & 200 & 10 (fixed)
\end{tabular}
\end{table}

@@

\begin{figure}[ht!]
  \centering
  \includegraphics[width=0.8\textwidth]{figures/wad_fo_partitioning_410}
  \caption{Modelled H$_2$O partition coefficient between forsterite, wadsleyite and melt at 14 GPa, approximately the pressure of the 410 km discontinuity. Note the separate scales for solid/solid and solid/melt partitioning. The difference in melt compositions between this study and that of \cite{DDFK2005} results in much lower $D^{wad/melt}$ coefficients. Error bars on the wadsleyite datapoints are those reported by \cite{DDFK2005}.}
  \label{fig:partitioning_wad_fo}
\end{figure}

\begin{figure}[ht!]
  \centering
  \includegraphics[width=0.8\textwidth]{figures/rw_wad_partitioning_520}
  \caption{Modelled H$_2$O partition coefficient between wadsleyite, ringwoodite and melt at 18 GPa, approximately the pressure of the 520 km discontinuity. Note the separate scales for solid/solid and solid/melt partitioning. Error bars on the ringwoodite data points indicate the range of values reported by \cite{OMY2000}. Error bars on the wadsleyite data points indicate 1 s.d. uncertainties as reported by \cite{LSOK2011}.}
  \label{fig:partitioning_rw_wad}
\end{figure}

@@

>>>>

%% Group authors per affiliation:
\author{R. Myhill}
\address{Bayerisches Geoinstitut, Universit\"{a}t Bayreuth, Universit\"{a}tsstrasse 30, 95447 Bayreuth, Germany \\ Now at School of Earth Sciences, University of Bristol, Wills Memorial Building, Queens Road, Bristol BS8 1RJ, United Kingdom}
\cortext[mycorrespondingauthor]{Corresponding author: R. Myhill}
\ead{bob.myhill@bristol.ac.uk}

@@

Here we use the approach of bracketing the liquidus surface to determine melt composition as a function of temperature in the system MgO-SiO$_2$-H$_2$O (MSH). This investigation is focussed on the behaviour at 13 GPa, which corresponds to the pressure just above the mantle transition zone, and is therefore relevant to the release of hydroxyl-bearing melts during deep subduction and during upwelling of hydrous wadsleyite. We use these results to develop thermodynamic models for water solubility in melts, and apply these to investigate water partitioning between melts and the olivine polymorphs (olivine, wadsleyite and ringwoodite). We then use our results to discuss the composition and properties of hydrous melts under transition zone conditions.

\section{Experimental and analytical methods}
Constraining melt compositions remains a major challenge for the field of high-pressure experimental petrology. The primary problem is that water-rich melts are unquenchable. During quenching, some of the water in the melt crystallises as hydrous minerals such as brucite, but a significant proportion crystallises as water ice at high pressure, which is then lost as water vapour during decompression and subsequent sample preparation. The original H$_2$O contents of the partial hydrous melts are normally estimated by mass balance and/or by deficits in electron probe totals \citep[e.g.][]{OMY2000, DDFK2005, LSOK2011, MSUP2007}. Using mass balance to calculate H$_2$O contents depends on being able to estimate the volume proportions of solid and quench phases, recognise and subtract quench overgrowths where they exist, and calculate the effective ambient-pressure density of the melt. Such estimates are typically associated with significant uncertainties. Deficits in electron probe totals are sometimes reasonable for estimating water contents in solid phases, provided the electron probe standards are suitable, and the electron probe is exceptionally well-calibrated (even a small deficit implies a large water content). However, for quench melts the deficits are attributable to water held in quench phases \emph{and} porosity within the volume sampled by the defocused beam. For this reason, defocused beam techniques should probably be restricted to estimating non-volatile element ratios (such as Mg:Si).

We avoid the above pitfalls by instead bracketting the liquidus surface with a series of distinct starting compositions. This technique is not without its own challenges. For example, the composition must be well-constrained and not change during the experiment, which requires techniques to effectively minimise water loss from the capsule. It must also be possible to recognise the difference between liquidus and quench phases. Starting compositions were created from a mixture of high-purity brucite, quartz and platinic acid (hexahydroxyplatinate(IV); H$_2$Pt(OH)$_6$). We use platinic acid to increase water contents to those expected of hydrous melts in equilibrium with mantle phases under subduction/convecting mantle conditions. Platinic acid is a pale yellow compound which is stable to about 130 $^{\circ}$C \citep{Nagano2002}. Its relatively high stability, high water content and the inert nature of the breakdown product (in a system where redox reactions are absent) make it an excellent water source for high pressure hydrous melting experiments. Before weighing, quartz was dried at 1000 $^{\circ}$C for 12 hours, while brucite was heated to 250 $^{\circ}$C, also over 12 hours. Both powders were stored in a desiccating oven at 130 $^{\circ}$C. Platinic acid is hygroscopic, so was stored in a vacuum desiccator. Powders were weighed and ground dry in an agate mortar for 30 minutes, using a mask, goggles and gloves to avoid physical contact with the platinic acid. Prepared powders were stored in glass vials in a vacuum desiccator until use.

@@

\begin{figure}[ht!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/periclase_edited}
  \caption{The periclase-water phase diagram at 13 GPa, based on current experimental results. Model fits to the data use the melt formulation of \cite{SS1985}, with r=1 and K=K(T) (see discussion in main text). Properties of the anhydrous solid and melt are taken from published literature \citep{SLB2011, DKS2013}. An additional subregular interaction term is required to match the composition and temperature of brucite dehydration.}
  \label{fig:MH}
\end{figure}

@@

\begin{figure}[ht!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/forsterite_edited}
      \caption{The forsterite-water phase diagram at 13 GPa, based on current experimental results. The thermodynamics of pure forsterite and the anhydrous melt are taken from the literature \citep{SLB2011, DKS2013}. At high temperatures, the forsterite liquidus is metastable with respect to periclase + melt. High temperature data come from the study of \cite{NDMPMF2016}.}
  \label{fig:foH}
\end{figure}
\clearpage

@@

\begin{figure}[ht!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/enstatite_edited}
  \caption{The clinoenstatite-water phase diagram at 13 GPa, based on current experimental results. The thermodynamics of pure high pressure clinoenstatite and the anhydrous melt are taken from the literature \citep{SLB2011, DKS2013}. High temperature data come from the study of \cite{NDMPMF2016}.}
  \label{fig:eoH}
\end{figure}

@@

\begin{figure}[ht!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/stishovite_edited}
      \caption{The silica-water phase diagram at 13 GPa, based on current experimental results. The thermodynamics of stishovite and the anhydrous melt are described in the text. At high temperatures, the stishovite liquidus is metastable with respect to coesite and melt.}
  \label{fig:SH}
\end{figure}

@@

@@

\begin{figure}[ht!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/experimental-ternary}
  \caption{A preliminary ternary liquidus diagram at 13 GPa, based on current experimental results. Contours are in $^{\circ}$C, and mineral abbreviations are liquidus fields as follows: per - periclase, fo - forsterite, en - clinoenstatite, stv - stishovite, coe - coesite, phB - phase B, br - brucite, chond - hydroxychondrodite, ice - ice VII. The locations of the contours are constrained by observed liquidus phases and the liquidus-bracketing results presented here. Previous work on hydrous phase stability and Mg:Si ratios for cotectic melts are also included \citep{MSUP2007}. The MgO-SiO$_2$ binary is constructed using a Margules solution model for the melt (see Supplementary Material).}
  \label{fig:ternary}
\end{figure}

@@

@@

Partitioning of hydrogen between mineral and melt phases can be described using a partition coefficient D:
\begin{equation}
  D^{a/b}_{H_2O} = c^a_{H_2O} / c^b_{H_2O}
\end{equation}
where $c^i_{H_2O}$ is the concentration of H$_2$O (in wt\%) in phase $i$ \citep{KB2006}. In quenchable phases, there are a number of ways that concentration can be measured, including FTIR, SIMS and ERDA. In the case of melts, which are unquenchable, $c^{melt}_{H_2O}$ in partitioning studies has most commonly been estimated by mass balance or by assuming that the deficit in electron probe totals is due to the fraction of oxygen bonded to hydrogen. In two studies on Fe-free wadsleyite, \cite{DDFK2005} and \cite{LSOK2011} present estimates of water content in melts calculated from electron probe deficits. A similar study on ringwoodite was undertaken by \cite{OMY2000}. The T-X$_{H_2O}$ dependencies of the melts in equilibrium with wadsleyite or ringwoodite are strikingly different (Figure \ref{fig:fo_wad_rw_melt}). For example, melt in equilibrium with wadsleyite at 15 GPa and 1400 $^{\circ}$C was estimated to contain 10.6-13.3 wt\% H$_2$O, while melt compositions in equilibrium with ringwoodite at 20 GPa and 1400 $^{\circ}$C suggest water contents of 57-66 wt\% H$_2$O. Modelled Mg$_2$SiO$_4$ melting temperatures increase by only 40 K between 15 and 20 GPa, and entropies of melting at these two pressures are not very different (92.4 vs. 97 J/K/mol), so unless the thermodynamics of mixing in the liquid change markedly between these pressures, estimates of $c^{melt}_{H_2O}$ for the ringwoodite or wadsleyite studies (or both) must be somewhat inaccurate. We note that two of the three melt compositions in equilibrium with ringwoodite fall above the temperatures indicating that hydrogen exists only as molecular H$_2$O \citep{SS1985}. In addition, the calculated liquidus temperatures in the wadsleyite studies at 15 GPa are lower than those reported for olivine at 5.5 GPa \citep{Inoue1994}. Both of these observations, therefore, raise inconsistencies which are difficult to address within a self-consistent thermodynamic model.

@@

\begin{figure}[ht!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/fo_wad_rw_edited}
  \caption{Water concentrations in wadsleyite, ringwoodite and melt, modelled at 15 and 20 GPa. Experimental data is taken from three published studies \citep{OMY2000,DDFK2005,LSOK2011}. Unfilled and filled symbols correspond respectively to the estimates of solid and melt H$_2$O contents.}
  \label{fig:fo_wad_rw_melt}
\end{figure}

@@

\begin{table}[]
\centering
\caption{Thermodynamic models for water solubility in forsterite, wadsleyite and ringwoodite. The forsterite model is taken from the literature \cite{ZGK2004}.}
\label{table:partitioning}
\begin{tabular}{lllll}
Phase & $P_{\textit{\tiny{ref}}}$ (GPa) & A [$10^{-12}$ kg/(kgPa)] & $\Delta H_{P_{\textit{\tiny{ref}}}}$ [kJ/mol] & $\Delta V$ [cm$^3$/mol] \\
\hline
Olivine & 12 & 13.5 & 170 & 10 \\
Wadsleyite & 15 & 1.8 & 160 & 10 \\
Ringwoodite & 20 & 1.2 & 200 & 10 (fixed)
\end{tabular}
\end{table}

@@

\begin{figure}[ht!]
  \centering
  \includegraphics[width=0.8\textwidth]{figures/wad_fo_partitioning_410_edited}
  \caption{Modelled H$_2$O partition coefficient between forsterite, wadsleyite and melt at 14 GPa, approximately the pressure of the 410 km discontinuity. Note the separate scales for solid/solid and solid/melt partitioning. The higher melt hydroxyl contents in this study compared to that of \cite{DDFK2005} results in much lower $D^{wad/melt}$ coefficients. Error bars on the wadsleyite datapoints are those reported by \cite{DDFK2005}.}
  \label{fig:partitioning_wad_fo}
\end{figure}

\begin{figure}[ht!]
  \centering
  \includegraphics[width=0.8\textwidth]{figures/rw_wad_partitioning_520_edited}
  \caption{Modelled H$_2$O partition coefficient between wadsleyite, ringwoodite and melt at 18 GPa, approximately the pressure of the 520 km discontinuity. Note the separate scales for solid/solid and solid/melt partitioning. Error bars on the ringwoodite data points indicate the range of values reported by \cite{OMY2000}. Error bars on the wadsleyite data points indicate 1 s.d. uncertainties as reported by \cite{LSOK2011}.}
  \label{fig:partitioning_rw_wad}
\end{figure}

@@

The authors thank Raj Dasgupta and Mike Walter for their insightful comments which greatly improved the manuscript.
